% Part: first-order-logic
% Chapter: syntax-and-semantics
% Section: Semantic Notions

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{fol}{syn}{sem}
\olsection{Semantic Notions}

\begin{explain}
Give the definition of !!{structure}s for first-order languages, we can
define some basic semantic properties of and relationships between
sentences.  The simplest of these is the notion of \emph{validity} of
a sentence.  A sentence is valid if it is satisfied in every
!!{structure}.  Valid sentences are those that are satisfied regardless of
how the non-logical symbols in it are interpreted.  Valid sentences
are therefore also called \emph{logical truths}---they are true, i.e.,
satisfied, in any !!{structure} and hence their truth depends only on the
logical symbols occurring in them and their syntactic !!{structure}, but not
on the non-logical symbols or their interpretation.
\end{explain}

\begin{defn}[Validity]
A sentence $!A$ is \emph{valid}, $\Entails !A$, iff $\Sat{M}{!A}$ for every
!!{structure}~$\Struct M$.
\end{defn}

\begin{defn}[Entailment]
A set of sentences~$\Gamma$ \emph{entails} a sentence~$!A$, $\Gamma \Entails 
!A$, iff $\Sat {M}{\Gamma}[s]$ implies $\Sat{M}{!A}[s]$ for every !!{structure}
~$\Struct M$ and variable assignment $s$.
\end{defn}


\begin{defn}[Satisfiability]
A set of sentences~$\Gamma$ is \emph{satisfiable} if $\Sat{M}{\Gamma}[s]$ for 
some !!{structure}~$\Struct M$ and some variable assignment $s$, i.e. $\Gamma 
\nvDash \lfalse$.  If $\Gamma$ is not satisfiable it is called \emph{ 
unsatisfiable}, i.e. $\Gamma \Entails \lfalse$.
\end{defn}

\begin{prop}
A sentence $!A$ is valid iff $\Gamma \Entails !A$ for every set of
sentences~$\Gamma$.
\end{prop}

\begin{proof}
For the forward direction, let $!A$ be valid, and let $\Gamma$ be a set of 
sentences. Let $\Struct M$, $s$ be so that $\Sat{M}{\Gamma}[s]$. Since $!A$ 
is valid, $\Sat{M}{!A}[s]$, hence $\Gamma \Entails !A$.

For the contrapositive of the reverse direction, let $!A$ be invalid, so 
there is $\Struct M$, $s$ with $\Sat/{M}{!A}[s]$. When $\Gamma = \{ \ltrue \}$
  , since $\ltrue$ is valid, $\Sat{M}{\Gamma}[s]$. Hence, there is a $\Struct 
M $ so that $\Sat{M}{\Gamma}[s]$ but $\Sat/{M}{!A}[s]$ for a variable 
assignment $s$, hence $\Gamma$ does not entail $!A$.
\end{proof}

\begin{prop}
\ollabel{prop:entails-unsat}
$\Gamma \Entails !A$ iff $\Gamma \cup \{\lnot !A\}$ is unsatisfiable.
\end{prop}

\begin{proof}
For the forward direction, suppose $\Gamma \Entails !A$ and suppose to the 
contrary that there is $\Struct M$, $s$ so that $\Sat{M}{\Gamma   \cup \{ 
\lnot !A \}}[s]$. Since $\Sat{M}{\Gamma}[s]$ and $\Gamma \Entails !A$, $\Sat{M
}{!A}[s]$. Also, since $\Sat{M}{\Gamma\cup \{ \lnot !A \}}[s]$, $\Sat{M}{\lnot
 !A}[s]$, so we have both $\Sat{M}{!A}[s]$ and $\Sat/{M}{!A}[s]$, a 
contradiction. Hence, there can be no such !!{structure} $\Struct M$, so $
\Gamma \cup \{ !A \}$ is unsatisfiable.

For the reverse direction, suppose $\Gamma \cup \{ \lnot !A \}$ is 
unsatisfiable. So for every $\Struct M$, $s$, either $\Sat/{M}{\Gamma}[s]$ or 
$\Sat/{M}{\lnot !A}[s]$. Hence, for every $\Struct M$, $s$ with $\Sat{M}{
\Gamma}[s]$, $\Sat{M}{!A}[s]$, so $\Gamma \Entails !A$.
\end{proof}

\begin{prob}
\begin{enumerate}
\item Show that $\Gamma \Entails \bot$ iff $\Gamma$ is unsatisfiable.
\item Show that $\Gamma, !A \Entails \bot$ iff $\Gamma \Entails \lnot !A$.
\item Suppose $c$ does not occur in $!A$ or $\Gamma$.  Show that
  $\Gamma \Entails \lforall[x][!A]$ iff $\Gamma \Entails
  \Subst{!A}{c}{x}$.
\end{enumerate}
\end{prob}


\begin{prop}
If $\Gamma \subseteq \Gamma'$ and $\Gamma \Entails !A$, then $\Gamma'
\Entails !A$.
\end{prop}

\begin{proof}
Suppose that $\Gamma \subseteq \Gamma'$ and $\Gamma \Entails !A$. Let $\Struct
  M$, $s$ be such that $\Sat{M}{\Gamma'}[s]$; then $\Sat{M}{\Gamma}[s]$, and 
since $\Gamma \Entails !A$, we get that $\Sat{M}{!A}[s]$. Hence, whenever $
\Sat{M}{\Gamma'}[s]$, $\Sat{M}{!A}[s]$, so $\Gamma' \Entails !A$.
\end{proof}


\begin{thm}[Semantic Deduction Theorem]
$\Gamma \cup \{!A\} \Entails !B$ iff $\Gamma \Entails !A \lif !B$.
\end{thm}

\begin{proof}
For the forward direction, let $\Gamma \cup \{ !A \} \Entails !B$ and let $
\Struct M$, $s$ be so that $\Sat{M}{\Gamma}[s]$. If $\Sat{M}{!A}[s]$, then $
\Sat{M}{\Gamma \cup \{ !A \}}[s]$, so since $\Gamma \cup \{ !A \}$ entails $!B
$, we get $\Sat{M}{!B}[s]$. Therefore, $\Sat{M}{!A \lif !B}[s]$, so $\Gamma 
\Entails !A \lif !B$.

For the reverse direction, let $\Gamma \Entails !A \lif !B$ and $\Struct M$, $
s$ be so that $\Sat{M}{\Gamma \cup \{ !A   \}}[s]$. Then $\Sat{M}{\Gamma}[s]$
, so $\Sat{M}{!A \lif !B}[s]$, and since $\Sat{M}{!A}[s]$, $\Sat{M}{!B}[s]$. 
Hence, whenever $\Sat{M}{\Gamma \cup \{   !A \} }[s]$, $\Sat{M}{!B}[s]$, so $
\Gamma \cup \{ !A \} \Entails !B$.
\end{proof}


\end{document}
